---
title: Tidy Data
author: "30 minutes"
output:
    ioslides_presentation
---

## Data Shapes - short versus long {.smaller}

### Short

```
animal    nToes   size                smell
chicken   4       small               so-so
cow       2       big         objectionable
pig       2       medium  more_objectionable
```
### Long

```
animal   variable              value
cow     size                	big
cow    	nToes                     2
cow    	smell         objectionable
pig     size                 medium
pig    	nToes                     2
pig    	smell    more_objectionable
chicken size                  small
chicken nToes                     4
chicken smell                 so-so
```

## But the long form is stupid! {.build}

<h2>Why would I ever use it??</h2>

## I'm glad you asked. {.build}

### Two reasons come up most often:

> * You want to take advantage of `group_by()` in the `dplyr` package to do complex stuff

> * You want to take advantage of `facets` in `ggplot2`

## dplyr and ggplot2 just work better on the long form 

(which makes sense, as all these packages are part of the `tidyverse`)

## Example data {.smaller}

```{r message=FALSE} 
filepath <- "http://hompal-stats.wabarr.com/datasets/studentgrades.txt"
grades <- read.table(filepath, header=T, sep=',')
head(grades)
```

## Two main functions {.build}

**`gather()`**  takes multiple columns, and collapse them into two key - value columns 

**`spread()`**  takes two columns (key - value pairs) and spreads them into separate columns

## gather() {.smaller}

Collapses multiple columns into two key - value pairs


```{r message=FALSE}
library(tidyverse)
head(grades)
grades <- gather(grades, key = "test", value="grade", exam1, exam2, exam3)
head(grades)
```

This allows you to use the `test` variable in facets, to make multiple plots easily

## gather() {.smaller}

```{r echo=TRUE, fig.width=8, fig.height=4}
qplot(x=TA, y=grade, geom='boxplot', data=grades, fill=TA) + 
  facet_wrap(~test) + 
  theme_bw(15) + 
  theme(axis.title.x=element_blank(),
        axis.text.x=element_blank(),
        axis.ticks.x=element_blank())
```


## spread() {.smaller}

Does the opposite of `gather()`. Spreads single column across multiple columns

```{r}
head(grades)
grades %>% spread(key=test, value=grade)
```

## Challenge -  using this dataset {.smaller}

```{r message=FALSE}
filepath <- "http://hompal-stats.wabarr.com/datasets/barr_astrag_2014.txt"
astrag <- read.table(filepath, header=TRUE, sep="\t")
head(astrag)
```

## Challenge - Make this Plot {.smaller}

```{r echo=FALSE, fig.width=8.5, warning=FALSE}
gAstrag <- astrag %>% 
           gather(key="measurement", 
                  value="value", ACF:WAT)
qplot(x=Habitat, y=value, data=na.omit(gAstrag), 
      fill= Habitat, geom="boxplot") + 
      facet_wrap(facets=~measurement, 
                 scales="free_y") + theme_bw()
```