---
title: "Homework #1 - Solution"
output: html_document
---

## 1.  Indexing Vectors

### 5 points possible

*  1pt  length
*  1pt  first four
*  1pt  last 8
*  1pt  various
*  1pt  reverse

```{r}
nouns <- c("apple", "flower", "insect", "lettuce", "knife", "dog", "cloud", "person", "cabinet", "flower" )

#length
length(nouns)

#first four
nouns[1:4]

#last 8
nouns[3:10]

#various elements
nouns[c(1, 3:6, 10)]

#reverse
nouns[10:1]

#alternate reverse
rev(nouns)

```


## 2. Using functions  

### 7 points possible

*  1pt  grades
*  1pt  curve
*  1pt  sd
*  1pt  min
*  1pt  max
*  1pt  mean
*  1pt  hist

```{r}
grades <- rnorm(200, mean = 68, sd=10)

#curve
curvedgrades <- grades + 7
curvedgrades

sd(curvedgrades)

min(curvedgrades)

max(curvedgrades)

mean(curvedgrades)

hist(curvedgrades)
```
    
## 3. Organizing data

### 4 points possible

*  2pts - fix data
*  2pts - describe problems 

The original data file had duplicated column names (length and diameter), relying on another column to differentiate the anatomical element.  This is a no, no.  Column names must stand alone (e.g. tibia_diameter).

The rows containing species names violate the 1 row = 1 observation priciple.  These rows should be turned into a column which records the species for each observation. 

## 4. Manipulation

### 7 pts possible

*  1pt  gorilla
*  2pts  sex
*  1pt  table
*  1pt  hist
*  1pt  small
*  1pt  big

```{r}
gorilla <- read.table("http://hompal-stats.wabarr.com/datasets/gorilla_sizes.txt", header=TRUE)

#figure out sex

#make a new column for sex, with 59 NA values as a placeholder
gorilla$sex <- rep(NA, 59)

#replace the NA values with 'Male' or 'Female' with the help of grep()
gorilla$sex[grep("m", gorilla$specimen)] <- "Male"
gorilla$sex[grep("f", gorilla$specimen)] <- "Female"

table(gorilla$sex)

hist(log(gorilla$skull))


smalls <- gorilla[gorilla$skull > 250,]
smalls

bigs <- subset(gorilla, subset = skull>250)
bigs
```
    
    