---
title: "Homework # 7 Solution"
output: html_document
---

## 6 pts total

```{r}
bovids <- read.table("http://hompal-stats.wabarr.com/datasets/bovid_occurrences.txt", header=TRUE, sep="\t")
head(bovids)
```

## Problem 1 - 2 pts
```{r}
library(tidyr)
spread_bovids <- spread(bovids, key=site, value=count)

#convert the taxon column to row.names instead, as these will be preserved in the matrix
#that we later use for the analysis
row.names(spread_bovids) <- spread_bovids$taxon
spread_bovids<-spread_bovids[,2:9]

```
## Problem 2 - 2pts

```{r fig.width=12, fig.height=12}
library(ggplot2)
ggplot(aes(x=taxon, y=count, fill=taxon), data=bovids) + 
  geom_bar(stat = "identity") + 
  facet_wrap(~site) + 
  theme_bw(30) + 
  theme(axis.text.x = element_text(angle = 45, hjust = 1))
```

## Problem 3 - 2pts

### Chi square is a good choice here as all the expected cell frequencies are well above 5.  Thus, it can be used to test the hypothesis that row and column variables are associated. 


```{r}
myTest <- chisq.test(spread_bovids)
```


### Just to make sure, lets check that our expected values are high enough.

```{r}
myTest$expected
```


### Now, we look at the results of the test

```{r}
myTest
```

### There is strong evidence that site and taxon are associated.  The null hypothesis is that there is no association between site and taxon.  Our p value from the chi square analysis is vanishingly small, which means that data as extreme as ours are extremely unlikely assuming the null hypothesis. 