---
title: "Basics"
---

## The Interactive Interpreter

Open R and type `2 + 2`.  R tells you that the answer is `4`. Congrats!  You are now using R. This is the most basic way to interact with the interpreter.  Later we will see how you can save a complicated series of commands to a script file and excecute them automatically. 

## Assignment

Sometimes it is useful to assign the results of computations to a named variable that we will use later.  We can create a named variable using the `<-` assignment operator.  For instance, we can do the following.

```{r}
myVariable <- 2 + 2
```

Now the result from this computation is assigned to `myVariable` and saved for later use.  Later, we can call this  variable like so: 

```{r}
myVariable * myVariable
```

## Vectors

A vector is a series of values stored in a single object.  You can create a vector using the `c()` function (**c** stands for **combine**). **Note**: you can only store a single type of data in a vector (e.g. numeric or character data).

```{r}
vector1 <- c(1,2,3,4,5,6,7,8,9,10)
vector2 <- c(11,12,13,14,15,16,17,18,19,20)
```

Vectors are very powerful, because many of the basic functions and operations in R are **vectorized**, meaning that they operate on all elements of a vector in sequence.  For example, to add the elements of `vector1` with the corresponding elements in `vector2`, you simply add the vectors, because the `+` function is vectorized. You can assume that **most** of the base functions in R are vectorized. This is one of the fundamental advantages of R. 

```{r}
vector1 + vector2
```

### Accessing values within a vector

Vectors can be **indexed** with brackets `[]` to get a subset of values.  What value do you get by typing `vector2[3]` in the interpreter? 

### <span class="mega-octicon octicon-puzzle"></span> Challenge <span class="mega-octicon octicon-puzzle">

Explain to the person sitting next to you how this works. HINT: R evaluates expressions 'from the inside out', with regards to brackets and parentheses.

```{r}
vector2[c(1,4,5)]
```

## Data Frames

Data frames are a convenient way of storing multiple related vectors of data together in a single object.  They are analogous to a spreadsheet where each row corresponds to an individual (e.g., specimen, species) and each row corresponds to some observation about that individual.

### Accessing data in a dataframe

Using the built in dataframe `iris`, you can explore how to access data within a dataframe.  The most common ways are with the `$` syntax and with the `[row,column]` syntax. `$` is used to access a named column within an dataframe.  The `[row, column]` syntax is used to identify the index number of the row and column desired.  Note that leaving a blank indicates ALL rows or columns.  So `iris$Species` is the same as `iris[,5]`.

```{r}
#first few rows of iris
head(iris)

#get the Sepal.Length column
iris$Sepal.Length

#get the value in the first row, third column
iris[1,3]

#get the whole first row
iris[1, ]
```

## Functions

Functions are the heart of the R language. A function is just a series of commands that is assigned a name.  Functions accept **arguments**, perform a series of commands using the argument values, then **return** a single object. Many functions have default values for arguments. You can see all the arguments and default values by using the `?` operator like this `?mean`.

Notice that the value for `na.rm = FALSE`.  This means that by default, `NA` values are not removed, and if you have some `NA` val

