---
title: "Distributions"
output: html_document
---

## Random Variables

A ***random variable*** --in statistical terms-- is a variable whose value depends on random chance. Each random variable has one or more parameters governing the probability of different outcomes. There are two types of random variables:  ***discrete random variables*** which have a limited number of possible discrete outcomes and ***continuous random variables*** which have a (theoretically) unlimited number of possible outcomes. 

### <span class="mega-octicon octicon-puzzle"></span> Challenge
Think of two examples of a discrete random variables and two examples of a continuous random variables for your area of scientific interest. What is the sample space for the random variable?

## Probability Distributions

If we plot the values from a random variable using a histogram, we call the result a ***probability distribution***.  The various distributions form the basis of parametric statistics. There are four main R functions associated with each probabibility distribution. In of the below, you will replace $dist$ with the abbreviation for that distribuction (norm, binom, etc.)

* rdist - random values drawn from the $dist$ function
* ddist - probability density of $dist$ at a particular point
* pdist - cumululative (tail) probability of the $dist$ fuction
* qdist - quantiles of $dist$ (quantiles are the converse of cumulative probability)

## Discrete Distributions

### Binomial Distribution

A **Bernouili trial** describes an event that has exactly two possibile outcomes: success and failure.  Success occurs with a given probability, for example $p = 0.2$. We know how likely a trial is to result in success, but any given trial may result in either success or failure, and this cannot be predicted for a single trial (the outcome is stochastic).

The result of a series of $n$ Bernoulili trials with $X$ successful outcomes results in a binomial random variable.

The expected value of the binomial distribution is $E(x) = np$

Binomial random variables can be simulated in R using the `rbinom()` function. 

```{r}
# 10 observations, 1 trial
rbinom(n = 10, size=1, prob = 0.5)

# 1 observation, 10 trials
rbinom(n = 1, size=10, prob = 0.5)

# 10 observations, 10 trials
rbinom(n = 10, size=10, prob = 0.5)

```

We can use the `dbinom()` function to get a sense of the shape of the binomial distrubution.

```{r}
library(ggplot2)
qplot(x=1:25, y=dbinom(1:25, 25, prob = 0.5), geom="line", main="binomial density prob=0.5")
qplot(x=1:25, y=dbinom(1:25, 25, prob = 0.2), geom="line", main="binomial density prob=0.5")
```

We can also look at this empirically by simulating data. The graph shows that the most common outcome of 5000 trials of a binomial function with a probability of 0.5 is to have approximately 2500 sucessful outcomes.  It is significantly less likely to get many more or less than 2500 successes.  


```{r}
qplot(rbinom(n = 5000, size=5000, prob = 0.5), main="binomial distribution p = 0.5")
```


### Poisson Distribution

Is similar to the binomial distribution, but describes rare events, when the number of trials $n$ is unknown. Requires a single rate parameter $\lambda$. The expected value and variance for the poisson distribution are both equal to lamba

```{r}
rpois(10, lambda = 0.2)

qplot(x=1:5, y=dpois(1:5, lambda = 0.2), main="Poisson distribution with lambda = 0.2", geom="line")
```

## Continuous Distributions

### Uniform Distribution

The uniform distribution represents a function in which the probability density is equal for each sub-interval across the a given range.  This results in a flat frequency distribution. The expected value over the range $a$ to $b$ is $(a + b)/2$

```{r}
runif(10, min = 1, max = 10)

qplot(runif(1000), min=1, max=10, main="uniform distribution from 1 to 10")
```

### Normal Distribution

The normal distribution (also known as the Gausian distribution) is the familiar bell-curve shaped distribution that is symetrical around the mean, with diminishing tails as you move away from the mean. Many phenomena in nature are distributed as a normal distribution.  The normal distribution has two parameters, the mean ($\mu$) and the standard deviation ($\sigma$).

```{r}
rnorm(10, mean=0, sd=1)

qplot(rnorm(10000), mean=0, sd=1, main="Standard normal distribution ")
```

### Log-normal Distribution

The log-normal distribution resembles a normal distribution when it is logged.  

```{r}
rlnorm(10, meanlog = 0, sdlog = 1)

lnorm <- rlnorm(1000, meanlog = 0, sdlog = 1)

qplot(lnorm, main="lognormal distribution")
qplot(log(lnorm), main="logged lognormal distribution")
```
